### 2023-08-04 paper/paper-papaja.Rmd
for submission

---
title: "Does mere exposure to ideas encourage belief in them?"
bibliography: references.bib
csl: apa7.csl
document-params: "10pt, letterpaper"

author-information: > 
    \author{{\large \bf Justin Mikell (jmikell@asu.edu)} \\ Arizona State University
    \AND {\large \bf Derek Powell (dmpowell@asu.edu)} \\ School of Social and Behavioral Sciences \\ Arizona State University}

abstract: >
    Numerous psychological findings have shown that mere exposure to ideas makes those ideas seem more true, a finding commonly referred to as the “illusory truth” effect [e.g. @hasher.etal1977]. In the presence of pervasive misinformation, this basic feature of cognition may undermine the functioning of a democratic society [@pennycook.etal2018]. However, genuine beliefs do not only produce judgments of truth, they also imply other beliefs and drive decision-making. Here, we sought to examine whether mere exposure to statements produces genuine beliefs by examining whether people draw inferences from statements after mere exposure. Surprisingly, and in contrast to familiarity-based accounts of the illusory truth effect [e.g. @dechene.etal2010], we found that exposure to “premise” statements affected participants’ truth ratings for novel “implied” statements. This “illusory implication” effect suggests that exposure to false statements has further-reaching impacts than previously thought and calls for a new mechanistic account of these effects.

keywords: >
    Illusory truth; Metacognition; Cognitive Psychology; Misinformation

output: 
  cogsci2016::cogsci_paper

final-submission: \cogscifinalcopy

# header-includes:
#   - \usepackage{float}
---

@@

```{r}
library(tidyverse)
library(patchwork)

@@

@@

@@

  res_table <- summary(model)$fixed

  res <- res_table[coef_name,] %>% 
    mutate_if(is.numeric, ~round(.x, digits=3))

  res_text <- paste0(res[[1,1]], ", 95% CI [", res[1,3],", ",res[1,4], "]")
  return(res_text)
}

# report_reg_coef(fit_imp_exp2, "familiarityTRUE")
```

@@

@@

exp1_age <- median(df_demo$age, na.rm=TRUE)
exp2_age <- median(df2_demo$age, na.rm=TRUE)
```

@@

The effect is one of "illusory" truth because it occurs following "mere exposure" to statements. That is, it occurs when the statements are seen or heard in a non-communicative context, such as when they are read during a true/false quiz. Generally, being told something---even by a complete stranger of unknown trustworthiness---is prima facie reason for believing it [@grice1989]. But reading a statement on a true/false quiz is not reason for believing it---the statement is being shown only to test the quiz-taker's knowledge, and is just as likely to be false as to be true.

There is something unsettling about the illusory truth effect and the lack of agency it implies over our own beliefs. Even more worrisome, there could be dire societal implications if merely being exposed to an idea causes people to adopt it as a belief: Pennycook and colleagues [-@pennycook.etal2018] argue that the illusory truth effect, combined with an environment of pervasive misinformation, has important implications for the functioning of democratic society. The illusory truth effect suggests that mere exposure to misinformation could have impacts that cannot be stopped by fact-checking labels [@pennycook.etal2018], nor effectively curbed by retractions or corrective information [e.g. @ecker.etal2011; @lewandowsky.etal2012].

However, there is more to _belief_ than ratings of truth. To illustrate, philosophers advancing dispositionalist theories of beliefs have focused on a number of observable behaviors that are indicative of belief [@schwitzgebel2021]. Though a person’s assent to a proposition (whether they agree with or judge it to be truthful) is an important marker and convenient measure of belief, other features of belief are just as essential. For instance, beliefs imply other beliefs: for instance, believing that “it is sunny out” implies the belief that “it is daytime”. And believing misinformation, such as “Hillary Clinton runs a sex-trafficking ring out of a pizza parlor” [see @robb2017] might imply the belief that “Hillary Clinton would make a bad president.” Beliefs also inform action and decision-making, and whether or not someone holds a belief can be judged (at least partly) from their decisions. When truly believed, misinformation can have drastic impacts. To illustrate, a recent study found that people who marked just one piece of vaccine misinformation as accurate were nearly twice as likely to be unvaccinated against COVID-19 compared to those who did not endorse any vaccine misinformation [@ognyanova.etal2021].

Bearing these features of beliefs in mind, the current state of illusory truth research does not demonstrate that mere exposure to a statement truly engenders _belief_ in that statement. Consider the mechanisms thought to underlie the illusory truth effect: most explanations attribute illusory truth effects to participants’ reliance on familiarity or fluency during the judgment process [@unkelbach2007; @wang.etal2016; @fazio.etal2015]. Roughly, when asked whether something is true or false, people search their memory for knowledge pertaining to belief in the statement, but they also rely on a meta-cognitive sense of familiarity. Familiarity is a sign that we have “heard this somewhere before.” According to foundational theories of communication, a general assumption that communicators strive to make true utterances is a prerequisite for successful communication and social functioning [@grice1989]. Thus, if having “heard something before” suggests that someone said it, then ecologically this is a reasonable cue to truth. Psychological studies that elicit the illusory truth effect hijack this meta-cognitive heuristic: They provide this sense of familiarity, but from a communicative context that lacks any reasonable assumption of truth.

This predominant theoretical account suggests that illusory truth effects do not engender genuine belief and may not have the deleterious real-world effects that some authors have feared [c.f. @pennycook.etal2018]. Applying this mechanistic account to the real-world, we might summarize the most likely process as follows: When people are exposed to a false statement they form a memory impression of it, but they do not draw inferences from this statement to revise other beliefs, nor do they plan future actions on its basis. In a real-world context, there would be essentially no impact from the exposure. In the experimental context however, the statement is seen again and a truth-judgment is requested. It is only at this time that the memory impression of the statement produces a feeling of familiarity and affects the truth judgments rendered at that time. Under this theoretical account, the impact of prior exposure depends on a subsequent re-exposure, coupled with a judgment or action to be taken at that time. This would make the real-world consequences of the illusory truth effect far narrower than some have argued.

Nevertheless, much as we might be relieved to dismiss them, concerns like those raised by Pennycook and colleagues [-@pennycook.etal2018] loom large enough to warrant further consideration. Could mere exposure to statements truly impact beliefs beyond providing a sense of familiarity? Could illusory truth effects instead reflect genuine belief?

Here, we sought to examine whether the illusory truth effect reflects genuine beliefs in statements following their mere exposure. One way to test this is to examine whether people draw inferences from statements after exposure to those statements in a context without any presupposition of truth. We designed a study to examine whether mere exposure to one statement (the “premise”) could affect truth ratings for another, different statement that it would logically imply (the “implication”). Such an “illusory implication” effect would demonstrate that mere-exposure has genuine impacts on beliefs beyond those explained by familiarity. In contrast, if the illusory truth effect is caused entirely by feelings of familiarity during the judgment process, then there should be no effect of prior exposure for new unfamiliar statements.

@@

Preregistration information for Experiment 1 can be found at https://osf.io/c4w8s/. Materials for both experiments can be found at https://osf.io/znq3y/.

@@

A total of 400 Participants were recruited through Amazon Mechanical Turk. Participants who failed basic attention check questions were excluded from analyses. These questions asked participants to simply give a particular response. For example, "This item tests if you are paying attention, please select 'Definitely False'." At the end of the study, participants were also asked if they had searched online to check any of the claims during the experiment (after being reassured that they would receive compensation regardless of their answers). Participants who failed any attention checks or who indicated they searched online were excluded from analysis. This left a final sample of `r nrow(df_demo)` (`r exp1_n_f` female, median age `r exp1_age` years-old).

```{r fig1, fig.env="figure", fig.pos = "h", fig.align = "center", fig.height=2.5, fig.width=3, fig.cap="Average truth ratings for Implication statements in Experiment 1, broken down by exposure (exposed vs. unexposed), instructions (fact vs quiz conditions), implication type (true-implied-false vs false-implied-true). For visualization purposes, means were calcaulated by translating ordinal responses onto a scale from -2.5 to 2.5. Error bars indicate standard errors. "}
fig1 <- df %>% 
  filter(phase=="test") %>% 
  filter(item_type == "B") %>% 
  group_by(exposed, exposure_cond, B_implied_true) %>% 
  summarize(
    M = mean(resp_num),
    se = sd(resp_num)/sqrt(n()),
    ll = M - se,
    ul = M + se
  ) %>% 
  mutate(
    exposure_cond = if_else(exposure_cond=="fact", "Fact condition", "Quiz condition"),
    exposed = if_else(exposed==1, "Exposed", "Unexposed"),
    B_implied_true = as.character(B_implied_true)
    ) %>% 
  ggplot(
    aes(x=B_implied_true, color=factor(exposed), shape=factor(exposed), y = M, ymin = ll, ymax=ul)
  ) +
  geom_pointrange(position=position_dodge(width=.5)) +
  # coord_flip() +
  scale_color_manual(values = c("#da1e28", "#0f62fe")) +
  labs(y = "Truth rating", x = "Implication type", color="Exposure", shape="Exposure") +
  facet_wrap(~exposure_cond) +
  theme_bw(base_size=9.5) +
  theme(panel.grid=element_blank(), legend.position="bottom") +
  theme(axis.text.x = element_text(angle = 10, vjust = 0.5, hjust=.5))

@@

```{r, include=F}
# library(kableExtra)
# kableExtra::kable(
#   tibble(
#     Type = c("True implied false", "False implied true"),
#     Premise = c(
#       "The Tour de France has been held every year since its inception in 1903 (False)",
#       "Henry Ford Invented the first automobile with the Model T (False)"
#     ),
#     Implied = c(
#       "The Tour de France was still held during WWI and WWII (False, implied True)",
#       "Karl Benz patented and sold the first automobile in 1886 (True, implied false)"
#     )
#   ),
#   caption = "Examples of premise and implied statement pairs"
# ) %>% 
#   column_spec(1, width = "4em") %>% 
#   column_spec(2, width = "10em") %>% 
#   column_spec(3, width = "10em")

@@

Participants first completed informed consent and a Captcha to limit bot participation.

The study proceeded in three phases: 1) an exposure phase, 2) a distraction phase, 3) and finally a testing phase. At the exposure phase, participants were presented with a subset of the "premise" statements (8 of the 24 total) as well as some control statements (12 true, 4 false). Then, after the distraction phase, the testing phase consisted of a true-or-false quiz. The items tested included both "premise" and "implied" statements from the 24 item-pairs.  For some of the tested "implied" statements, participants had previously seen the related "premise" statement (exposed implied test). For others, they had not seen the related statements (unexposed implied test). Similarly, they were tested on the "premise" statements they had previously seen during the exposure phase (exposed premise test) as well as new unseen premise statements (for which they had also not seen the related implied statements). Participants were randomly assigned into three counterbalancing conditions that varied which of the statement-pairs were assigned to each exposure/test combination.

At the beginning of the study, participants were randomly assigned to either the “fact” (_n_ = 100) or “quiz” (_n_ = 300) exposure condition and to one of three counterbalancing conditions. Participants assigned to the quiz and fact conditions received different instructions and performed different tasks in the exposure phase.

@@

Participants in the quiz condition were told that they were to be given a true/false quiz, so that some of the statements would be true and some false. These participants rated how confident they were that each of the presented statements were true or false. The purpose of the quiz condition was to provide "mere exposure" to these statements, to test for illusory truth and illusory implication effects. We tested for the illusory truth effect by comparing truth ratings for premise statements seen during exposure to those that had not been seen. Similarly, we tested for the illusory implication effect by comparing truth ratings for implication statements whose corresponding premise statements were and were not presented during the exposure phase. If mere exposure to "premise" statements affects participants' judgments of the "implied" statements, this would be evidence for an illusory implication effect.

@@

Then, participants advanced to the testing phase, where all participants were asked to rate their confidence that each statement was true or false. These ratings were made on a 6-point scale from "Definitely False" to "Definitely True". The test phase consisted of the 8 premise statements the participants previously saw, their respective matching 8 implication statements, 8 novel implication statements, 8 previously unseen premise statements, 12 true control statements, and 4 false control statements.

@@

## Results and discussion

Participants made their truth ratings in the test phase on a 6-point scale from “Definitely false” to “Definitely true”. To properly treat these Likert-style responses, the truth ratings were analyzed using multilevel Bayesian cumulative ordinal regression models [@burkner.vuorre2019] with random intercepts and slopes for participants and items. Cumulative ordinal regression models assume that participants’ discrete responses are driven by a continuous latent variable and a set of k-1 thresholds determining the range of the continuous variable corresponding to each of the ordinal response options. This model helps to account for potential differences in scale usage as well as the bounded nature of the response scale.

All models were fit using the brms R package [@burkner2017], with model posteriors estimated using the No-U-Turn Markov Chain Monte Carlo (MCMC) sampler implemented in Stan. Four MCMC chains were run for each model, with 2000 samples (1000 burn-in) drawn from each. Chains were assessed for convergence with $\hat{R}$ and the total estimated effective sample size was verified to be greater than 1000 for all parameters [@gelman.etal2014].

@@

It is quite appropriate that people should draw new inferences when they learn new facts. But would participants’ endorsements of the implication statements be affected by mere exposure to the premise statements in the “quiz” condition? To our surprise, it appears that they were. Figure 1 shows participants’ average truth ratings for the implication statements in the quiz condition. Participants gave higher truth ratings following exposure for false statements implied to be true, and somewhat lower truth ratings following exposure for true statements implied to be false.

@@

Experiment 2 was conducted to address the possibility that a pressure for internal consistency produced the effects on the implication-statement truth ratings observed in Experiment 1. Preregistration informationExperiment 2 can be found at https://osf.io/czkdh/.

>>>>

---
title             : "The title"
shorttitle        : "Title"

author: 
  - name          : "Justin Mikell"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "Arizona State University - West Campus, 4701 W. Thunderbird Rd., Glendale, AZ 85306"
    email         : "jmikell@asu.edu"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - "Investigation"
      - "Methodology"
      - "Visualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
  - name          : "Derek Powell"
    affiliation   : "1"
    role:
      - "Conceptualization"
      - "Formal Analysis"
      - "Methodology"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
      - "Validation"
      - "Supervision"

affiliation:
  - id            : "1"
    institution   : "Arizona State University"

authornote: |
  Arizona State University, School of Social and Behavioral Sciences

  Materials for all experiments can be found at https://osf.io/znq3y/. Preregistration information for Experiments can be found at https://osf.io/c4w8s/ (Exp. 1), https://osf.io/czkdh/ (Exp. 2), and https://osf.io/5ewa6 (Exp. 3).

abstract: |
  Numerous psychological findings have shown that mere exposure to ideas makes those ideas seem more true, a finding commonly referred to as the “illusory truth” effect [e.g. @hasher.etal1977]. In the presence of pervasive misinformation, this basic feature of cognition may undermine the functioning of a democratic society [@pennycook.etal2018]. However, genuine beliefs do not only affect judgments about individual statements, they also imply other beliefs and drive decision-making. Here, we sought to examine whether mere exposure to statements produces genuine beliefs by examining whether people draw inferences from statements after mere exposure. Surprisingly, and in contrast to fluency or familiarity-based accounts of the illusory truth effect [e.g. @dechene.etal2010], we found that exposure to “premise” statements affected participants’ truth ratings for novel “implied” statements. This “illusory implication” effect suggests that the consequences of exposure to false statements reach further than previously thought and calls for a new mechanistic account of these effects.

  <!-- https://tinyurl.com/ybremelq -->

keywords          : "Illusory truth; Metacognition; Cognitive Psychology; Misinformation"
wordcount         : "X"

bibliography      : "references.bib"

floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no

figurelist        : no
tablelist         : no
footnotelist      : no

classoption       : "man"
output            : papaja::apa6_pdf
---

```{r setup, include = FALSE}
library("papaja")
r_refs("r-references.bib")
```

```{r analysis-preferences}
# Seed for random number generation
set.seed(42)
# knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
```

@@

```{r}
library(tidyverse)
library(patchwork)
library(textstem)

@@

```{r fit-models-exp3, include=F}
df_all <- read_csv("../data/illusory-truth-main-2.csv")

df_passing <- df_all %>% 
  mutate(
    fail_checks = case_when(
      (item == "att_exp_titanic") & (response != "Statement B") ~ 1,
      (item == "att_exp_old") & (response != "Statement A") ~ 1,
      (item == "att_exp_wright") & (response != "Statement B") ~ 1,
      (item == "att_test_taxes") & (grepl("True", response)) ~ 1,
      (item == "att_test_pop") & (grepl("False", response)) ~ 1,
      (item == "att_test_tall") & (grepl("True", response)) ~ 1,
      instructions_check!="Mix of true and false" ~ 1,
      TRUE ~ 0
    )
  ) %>%
  group_by(subj_id) %>%
  filter(sum(fail_checks) < 1) %>% # remove anyone failing checks (none fail)
  filter(!grepl("att_", item)) %>% # (remove check items)
  select(-fail_checks) %>% 
  ungroup() %>% 
  mutate(
    suspicious = if_else(susp_check1 == "Yes", 1, 0)
  )

df3 <- df_passing  %>% 
  filter(phase=="test") %>% 
  mutate(
    response = ordered(response, levels = c("Definitely False", "Probably False", "Maybe False", "Maybe True", "Probably True", "Definitely True"))
  ) %>% 
  mutate(resp_num = recode(response,
                           `Definitely True` = 2.5,
                           `Probably True` = 1.5,
                           `Maybe True` = .5,
                           `Maybe False` = -.5,
                           `Probably False` = -1.5,
                           `Definitely False` = -2.5)
         ) %>% 
  mutate(
    B_implied_true = ifelse(item %in% c("war", "whitehouse", "gates", "jackson",
                                          "indonesia", "eiffel", "newyork", "presidents",
                                          "tour", "africa", "disney", "avengers"), 
                            "F implied T", "T implied F"),
    B_implied_true = relevel(factor(B_implied_true), ref = "T implied F"),
      acc_num = case_when(
        item_type=="B" & B_implied_true == "F implied T" ~ -1*resp_num,
        item_type=="B" & B_implied_true != "F implied T" ~ resp_num,
        item_type=="A" ~ -1*resp_num,
        item_type=="Falsehood" ~ -1*resp_num,
        TRUE ~ resp_num
        ),
    acc_ordered = ordered(acc_num)
  )

df_open_coded <- read_csv("../data/illusory-truth-main-2-free-responses-coded.csv") %>%
  select(subj_id, flagged)

df3 <- df3 %>% 
  left_join(df_open_coded, by = "subj_id")

## Main analyses

df_reg3 <- df3 %>% 
  mutate(
    familiarity = exposed,
    implication = case_when(
      B_implied_true == "F implied T" & exposed==1 ~ 1,
      B_implied_true == "T implied F" & exposed==1 ~ -1,
      TRUE ~ 0
    )
    # implied_true = if_else(B_implied_true == "F implied T" & exposure==1, 1, 0),
    # implied_false = if_else(B_implied_true != "F implied T" & exposure==1, 1, 0),
    ) %>% 
    filter(phase=="test", item_type=="B")

fit_imp_maximal <- brm(
  response ~ familiarity + implication + B_implied_true + (1 + familiarity + implication + B_implied_true |subj_id) + (1 + familiarity + implication|item),  # familiarity + implication + grouping (optional)
  data = df_reg3,
  family = cumulative(),
  chains = 4,
  cores = 4,
  iter = 3000,
  control = list(adapt_delta = .95),
  file="../local/fit_imp_rep_maximal"
)

# fit_imp <- brm(
#   response ~ familiarity + implication + B_implied_true + (1 |subj_id) + (1 |item),  # familiarity + implication + grouping (optional)
#   data = df_reg,
#   family = cumulative(),
#   chains = 4,
#   cores = 4,
#   file="local/fit_imp_rep"
# )

### excluding suspicious

df_reg3_nosus <- df_reg3 %>% 
    filter(suspicious==0)

fit_imp_maximal2 <- brm(
  response ~ familiarity + implication + B_implied_true + (1 + familiarity + implication + B_implied_true |subj_id) + (1 + familiarity + implication|item),  # familiarity + implication + grouping (optional)
  data = df_reg3_nosus,
  family = cumulative(),
  chains = 4,
  cores = 4,
  iter = 3000,
  control = list(adapt_delta = .95),
  file="../local/fit_imp_rep_maximal2"
)

df_reg3_noflag <- df_reg3 %>% 
    filter(suspicious==0, flagged==0)

fit_imp_maximal3 <- brm(
  response ~ familiarity + implication + B_implied_true + (1 + familiarity + implication + B_implied_true |subj_id) + (1 + familiarity + implication|item),  # familiarity + implication + grouping (optional)
  data = df_reg3_noflag,
  family = cumulative(),
  chains = 4,
  cores = 4,
  iter = 3000,
  control = list(adapt_delta = .95),
  file="../local/fit_imp_rep_maximal3"
)

```{r, item-fluency-analysis, include=F}
statements_raw <- read_csv("../items.csv")

clean_words <- function(x){
  lower <- str_to_lower(x)
  word_list <- str_split(lower, " ")
  lemmas <- lemmatize_strings(word_list[[1]])
  # no_stops <- lemmas[ !(lemmas %in% lemmatize_words(lexicon::sw_lucene)) ]
  lemma_stems <- stem_words(lemmas)

  out <- gsub("('|,)", "", lemma_stems)

  return(out)

}

statements <- statements_raw %>%
  mutate(
    prem_lst = map(premise, clean_words),
    imp_lst = map(implication, clean_words),
    overlap = map2_dbl(prem_lst, imp_lst, ~sum(.y %in% .x)), # how many words in implication were in premise?
    overlap_prop = overlap/length(imp_lst)
  )

statement_overlap <- statements %>%
  group_by(B_implied_true) %>%
  summarize(overlap = median(overlap_prop))

item_fluency_fig <- statements %>%
  mutate(B_implied_true = if_else(B_implied_true==1, "F implied T", "T implied F")) %>% 
  ggplot(aes(x = factor(B_implied_true), y = overlap_prop)) +
  stat_summary(fun.data=mean_se, fun.args = list(mult=2), 
                 geom="pointrange", color="red", shape=3, position=position_dodge()) +
  geom_jitter(width=.1, size=2) +
  # geom_dotplot(binaxis="y", stackdir="center", ,
  #              stackratio=1.5, dotsize=.6) +
  theme_bw(base_size = 12) +
  theme(panel.grid=element_blank()) +
  ylim(0,.5) +
  labs(y = "Prop. shared lemmas", x = "Item Group")
```

@@

mytable3 <- make_table_df(fit_imp_maximal) %>% 
  rename(`$CI_{2.5\\%}$` = `l-95% CI`,`$CI_{97.5\\%}$` = `u-95% CI`) %>%
  mutate(
    Term = case_when(
      Term == "B_implied_trueFimpliedT" ~ "implication type",
      # grepl("Intercept\\[", Term) ~ gsub("Intercept\\[(.*)\\]","$\\\\alpha_{\\1}$", Term),
      grepl("familiarity", Term) ~ "familiarity",
      # Term == "item_sd(Intercept)" ~ "$\\sigma_{item}$",
      # Term == "subj_sd(Intercept)" ~ "$\\sigma_{subj}$",
      TRUE ~ Term
    )
  ) %>% 
  make_table("Population coefficients of Bayesian regression model for Implication effects in Experiment 3.")

@@

  res_df <- broom.mixed::tidy(model, effects="fixed")
  res <- res_df %>% 
    filter(term == coef_name) %>% 
    mutate_if(is.numeric, ~round(., 3))

  res_text <- paste0(res[[1,4]], ", 95% CI [", res[[1,6]],", ",res[[1,7]], "]")
  return(res_text)
}

# report_reg_coef(fit_imp_exp2, "familiarityTRUE")
# report_reg_coef(fit_imp_fact, "exposed")
```

@@

df3_demo <- df3 %>% 
  group_by(subj_id) %>% 
  summarize(
    age = first(age),
    gender = first(gender)
  )

@@

exp3_n_f <- df3_demo %>% 
  filter(gender=="Female") %>% 
  nrow()

exp1_age <- median(df_demo$age, na.rm=TRUE)
exp2_age <- median(df2_demo$age, na.rm=TRUE)
exp3_age <- median(df3_demo$age, na.rm=TRUE)

exp3_susp_count <- df3 %>% 
  filter(suspicious==1) %>% 
  distinct(subj_id) %>% 
  nrow()

exp3_flagged_count <- df3 %>% 
  filter(suspicious==0) %>%
  distinct(subj_id, .keep_all = TRUE) %>% 
  filter(flagged==1) %>% 
  nrow()

## Relevance Statement

Three pre-registered experiments examined whether mere exposure to ideas might lead them to be adopted as beliefs. It is well-established that mere exposure, such as in a true-or-false quiz, can lead people to later rate statements as more true (the “illusory truth” effect). If people's beliefs are affected by mere exposure, then the pervasiveness of misinformation would be even more harmful than currently recognized. However, leading theories attribute illusory truth effects to feelings of familiarity or fluency rather than changes in beliefs. These metacognitive heuristics are unlikely to produce further inferences or drive decision-making, as would beliefs. In contrast, we demonstrate that mere exposure to statements affected participants' responses to other novel statements (what we term an “illusory implication” effect), suggesting that mere exposure can lead to genuine belief. Our findings illustrate new dangers in exposure to misinformation, showing that mere exposure may have more wide-ranging influences than previously understood.

\newpage

@@

The effect is one of "illusory" truth because it occurs following "mere exposure" to statements. That is, it occurs when the statements are seen or heard in a non-communicative context, such as when they are read during a true/false quiz. Generally, being told something---even by a source of unknown trustworthiness---is prima facie reason for believing it [@grice1989]. But reading a statement on a true/false quiz is not reason for believing it; the statement is just as likely to be false as to be true.

Why should mere exposure affect later ratings of truth? Predominant theoretical accounts attribute illusory truth effects to metacognitive experiences of fluency or familiarity [@dechene.etal2010]. Research on metacognition has shown that people's behavior and judgments are not only influenced by the content of their thoughts, but also by the phenomenological experience of processing those thoughts [For a review, see @schwarz2015].

Metacognitive processing "fluency" is the experienced ease of processing or thinking about information [@alter.oppenheimer2009; @wanke.hansen2015]. When thinking is easy and fluent, people tend to judge the targets of that thinking more favorably than if thinking is challenging and disfluent [@wanke.etal1997]. A number of findings have shown that processing fluency contributes to the illusory truth effect: Mere prior exposure to statements makes later processing of those statements more fluent, and this metacognitive experience of fluency leads people to rate those statements as more true [@difonzo.etal2016; @fazio.etal2015; @unkelbach2007; @wang.etal2016].

Other researchers have attributed the illusory truth effect to familiarity [e.g. @arkes.etal1991; @mitchell.etal2006]. Prior exposure to statements engenders a feeling of familiarity upon re-exposure to those statements. This metacognitive feeling of familiarity provides a cue that leads people to infer the statement is more likely to be true.

Both theoretical accounts provide a similar overall view of the illusory truth effect: Roughly, when asked whether something is true or false, people search their memory for knowledge pertaining to belief in the statement, but they also rely on their metacognitive experiences of familiarity and fluency. Familiarity and fluency are both signs that we have “heard this somewhere before.” According to foundational theories of communication, a general assumption that communicators strive to make true utterances is a prerequisite for successful communication and social functioning [@grice1989]. Thus, if having “heard something before” suggests that someone said it, then ecologically this is a reasonable cue to truth. Psychological studies that elicit the illusory truth effect hijack these metacognitive heuristics: They provide this sense of familiarity or fluency, but from a communicative context that lacks any reasonable assumption of truth.

There is something unsettling about the illusory truth effect and the lack of agency it implies over our own beliefs. Even more worrisome, there could be dire societal implications if merely being exposed to an idea causes people to adopt it as a belief: Pennycook and colleagues [-@pennycook.etal2018] argue that the illusory truth effect, combined with an environment of pervasive misinformation, has important implications for the functioning of democratic society. The illusory truth effect suggests that mere exposure to misinformation could have impacts that cannot be stopped by fact-checking labels [@pennycook.etal2018], nor effectively curbed by retractions or corrective information [c.f. @ecker.etal2011; @lewandowsky.etal2012].

However, there are more to _beliefs_ than ratings of truth. Though a person’s assent to a proposition (whether they agree with or judge it to be truthful) is an important marker and convenient measure of belief, other features of belief are just as essential. For one, beliefs imply other beliefs: for instance, believing that “it is sunny out” implies the belief that “it is daytime”. And believing misinformation, such as “Dominion voting machines were rigged in the 2020 election” might imply the belief that “Donald Trump actually won the 2020 election.” Beliefs also inform action and decision-making, and whether or not someone holds a belief can be judged (at least partly) from their decisions. When truly believed, misinformation can have serious consequences. For instance, a recent study found that people who marked just one piece of vaccine misinformation as accurate were nearly twice as likely to be unvaccinated against COVID-19 compared to those who did not endorse any vaccine misinformation [@ognyanova.etal2021].

Generally, it is these wider consequences of belief (beyond truth ratings for a specific proposition) that should be most concerning. If illusory truth effects operate only through metacognitive influences, then these effects may not have much impact outside the lab.  Under these theoretical accounts, the impact of prior exposure depends on a subsequent re-exposure to a now fluent or familiar statement. This would make the real-world consequences of the illusory truth effect far narrower than one might fear: These metacognitive mechanisms are liable to apply only in response to the exact statements previously seen, but are not as likely to generalize to other related ideas, or to drive actions or decision-making in novel contexts.

Nevertheless, concerns about how the influence of mere exposure might operate in contexts of pervasive misinformation [@difonzo.etal2016; @pennycook.etal2018] loom large enough to warrant further consideration. Could mere exposure to statements truly change beliefs?

A crucial feature of belief is that beliefs are connected with other related beliefs. Thus, one way to test whether mere exposure affects beliefs is to examine whether mere exposure to statements can affect truth ratings for other related statements.

A handful of studies are suggestive in this regard: Arkes and collagues [-@arkes.etal1991] found that mere exposure to one set of statements also affected judgments about novel related statements on the same topic (China). However, these findings could also be explained by a sense of familiary engendered by the shared topic of old and new statements [@arkes.etal1991].

Bacon [-@bacon1979] found that mere exposure to one statement ($P$) could lead to lower truth ratings for a new, contradictory statements($\neg P$). For instance, mere exposure to "crocodiles sleep with their eyes open" produced lower truth ratings for "crocodiles sleep with their eyes closed". It seems clear at least that familiarity cannot explain this pattern of results. However, other findings under very similar paradigms have partly contradicted these results [@garcia-marques.etal2015] and it is somewhat unclear whether fluency could account for these findings, i.e. whether contradictory statements can be expected to lead to fluent or disfluent processing.

Here, we sought to decisively examine whether mere exposure to statements could influence people's beliefs. To do so, we examined whether people draw inferences from statements after exposure in a context without any presupposition of truth. We designed a study to examine whether mere exposure to one statement (the “premise”) could affect truth ratings for a  different statement the premise would logically entail (the “implication”). Such an “illusory implication” effect would indicate that mere exposure has genuine impacts on beliefs extending beyond those attributable to metacognitive cues.

@@

@@

A total of 400 Participants were recruited through CloudResearch. Basic attention check questions were used to screen inattentive participants. These questions asked participants to simply give a particular response. Participants who failed any attention checks or who indicated during debrief that they had searched online for question answers were excluded from analysis. This left a final sample of `r nrow(df_demo)` (`r exp1_n_f` female, median age `r exp1_age` years-old).

```{r fig1, fig.env="figure", fig.pos = "h", fig.align = "center", fig.cap="Average truth ratings for Implication statements in Experiment 1, broken down by exposure (exposed vs. unexposed), instructions (fact vs quiz conditions), implication type (true-implied-false vs false-implied-true). For visualization purposes, means were calcaulated by translating ordinal responses onto a scale from -2.5 to 2.5. Error bars indicate standard errors. "}
fig1 <- df %>% 
  filter(phase=="test") %>% 
  filter(item_type == "B") %>% 
  group_by(exposed, exposure_cond, B_implied_true) %>% 
  summarize(
    M = mean(resp_num),
    se = sd(resp_num)/sqrt(n()),
    ll = M - se,
    ul = M + se
  ) %>% 
  mutate(
    exposure_cond = if_else(exposure_cond=="fact", "Fact condition", "Quiz condition"),
    exposed = if_else(exposed==1, "Exposed", "Unexposed"),
    B_implied_true = as.character(B_implied_true)
    ) %>% 
  ggplot(
    aes(x=B_implied_true, color=factor(exposed), shape=factor(exposed), y = M, ymin = ll, ymax=ul)
  ) +
  geom_pointrange(position=position_dodge(width=.5)) +
  # coord_flip() +
  scale_color_manual(values = c("#da1e28", "#0f62fe")) +
  labs(y = "Truth rating", x = "Implication type", color="Exposure", shape="Exposure") +
  facet_wrap(~exposure_cond) +
  theme_bw(base_size=9.5) +
  theme(panel.grid=element_blank(), legend.position="bottom") +
  theme(axis.text.x = element_text(angle = 10, vjust = 0.5, hjust=.5))

@@

@@

The study proceeded in three phases: 1) an exposure phase, 2) a distraction phase, and 3) a final testing phase.

At the beginning of the study, participants were randomly assigned to either the “fact” (_n_ = 100) or “quiz” (_n_ = 300) exposure condition and to one of three counterbalancing conditions. At the exposure phase, participants were presented with a subset of the "premise" statements (8 of the 24 total, counterbalanaced across participants) as well a set of control statements (12 true, 4 false). Participants assigned to the quiz and fact conditions received different instructions and performed different tasks in the exposure phase.

@@

Participants in the quiz condition were told that they were to be given a true/false quiz, so that some of the statements would be true and some false. These participants rated how confident they were that each of the presented statements were true or false. The purpose of the quiz condition was to provide "mere exposure" to these statements, to test for illusory truth and illusory implication effects.

@@

Finally, the testing phase for all participants consisted of a true-or-false quiz, where participants were asked to rate their confidence that each statement was true or false. These ratings were made on a 6-point scale from "Definitely False" to "Definitely True". The items tested included both "premise" and "implied" statements from the 24 item-pairs. Participants were quizzed on the 8 previously-exposed premise statements, as well as their corresponding "implied" statements. For comparison, they were also quizzed on 8 new premise statements and 8 unrelated implied statements. Participants were randomly assigned into three counterbalancing conditions that varied which of the statement-pairs were assigned to each exposure/test combination.

We tested for the illusory truth effect by comparing truth ratings for premise statements seen during exposure to those that had not been seen. Similarly, we tested for the illusory implication effect by comparing truth ratings for implication statements whose corresponding premise statements were and were not presented during the exposure phase. If mere exposure to "premise" statements affects participants' judgments of the "implied" statements, this would be evidence for an illusory implication effect.

@@

## Results

Participants made their truth ratings in the test phase on a 6-point scale from “Definitely false” to “Definitely true”. To properly treat these Likert-style responses, the truth ratings were analyzed using multilevel Bayesian cumulative ordinal regression models [@burkner.vuorre2019] with random intercepts and slopes for participants and items. Cumulative ordinal regression models assume that participants’ discrete responses are driven by a continuous latent variable and a set of $k-1$ thresholds determining the range of the continuous variable corresponding to each of the ordinal response options. This model helps to account for potential differences in scale usage as well as the bounded nature of the response scale.

All models were fit using the `brms` R package [@burkner2017], with model posteriors estimated using the No-U-Turn Markov Chain Monte Carlo (MCMC) sampler implemented in Stan. Four MCMC chains were run for each model, with 2000 samples (1000 burn-in) drawn from each. Chains were assessed for convergence with $\hat{R}$ and the total estimated effective sample size was verified to be greater than 1000 for all parameters [@gelman.etal2014].

@@

Figure 1 shows participants’ average truth ratings for the implication statements in the quiz condition. Participants gave higher truth ratings following exposure for false statements implied to be true, and somewhat lower truth ratings following exposure for true statements implied to be false.

@@

Experiment 2 was conducted to address the possibility that a pressure for internal consistency produced the effects on the implication-statement truth ratings observed in Experiment 1.
